---
title: 
author: 
output: pdf_document
---

\section{Model selection for the Gambia data}


The gambia data in the geoR package includes data for 1332 children in the Gambia. The binary response $Y_i$ s the indictor that child $i$ tested positive for malaria. Child $i$ lives in village $v_i\in \{1,…,65\}$. We use five covariates in $X_{ij}$.

Age: age of the child, in days

Netuse: indicator variable denoting whether (1) or not (0) the child regularly sleeps under a bed-net

Treated: indicator variable denoting whether (1) or not (0) the bed-net is treated (coded 0 if netuse=0)

Green: satellite-derived measure of the greenness of vegetation in the immediate vicinity of the village (arbitrary units)

PCH: indicator variable denoting the presence (1) or absence (0) of a health center in the village

We use the random effects logistic regression model

$$\mbox{logit}[\mbox{Prob}(Y_i=1)] = \alpha + \sum_{j=1}^pX_{ij}\beta_j + \theta_{v_i},$$

where  $\theta_v$ is the random effect for village $v$.  We compare three models for the village random effects via DIC and WAIC:

1. No random effects: $\theta_v=0$.

2. Gaussian random effects: $\theta_v\sim\mbox{Normal}(0,\tau^2)$.

3. Double-exponential random effects: $\theta_v\sim\mbox{DE}(0,\tau^2)$. 

\section{Load the data}

```{r}
library(geoR)
data(gambia)
Y <- gambia[,3]
X <- scale(gambia[,4:8])
s <- gambia[,1:2]
n <- length(Y)
p <- ncol(X)
# Compute the village ID
S <- unique(s) # Lat/long of the villages
m <- nrow(S)
village <- rep(0,n)
members <- rep(0,m)
for(j in 1:m){
d <- (s[,1]-S[j,1])^2 + (s[,2]-S[j,2])^2
village[d==0] <- j
members[j] <- sum(d==0)
}
size <- ifelse(members<25,1,2)
size <- ifelse(members>35,3,size)
table(size)

```


```{r}
pch  <- c(1,4,19) 

plot(gambia.borders, type="l", 
     asp=1,axes=F,cex.main=1.5,
     xlab="",ylab="",main="Village locations")
points(S[,1],S[,2],pch=pch[size])
legend("top",c("<25 children","25-35 children",">35 children"),pch=pch,cex=1.5,bty="n")
```

\section{Prep for JAGS}
```{r}
library(rjags)
   burn   <- 1000
   iters  <- 5000
   chains <- 2 
```

\section{Model 1: No random effects}

```{r}
mod <- textConnection("model{
     for(i in 1:n){
       Y[i]          ~ dbern(pi[i])
       logit(pi[i]) <- beta[1]        + X[i,1]*beta[2] + X[i,2]*beta[3] + 
                       X[i,3]*beta[4] + X[i,4]*beta[5] + X[i,5]*beta[6]
       like[i]      <- dbin(Y[i],pi[i],1) # For WAIC computation
     }
     for(j in 1:6){beta[j] ~ dnorm(0,0.01)}
   }")

   data   <- list(Y=Y,X=X,n=n)
   model  <- jags.model(mod,data = data, n.chains=chains,quiet=TRUE)
   update(model, burn, progress.bar="none")
   samps  <- coda.samples(model, variable.names=c("like"), 
                          n.iter=iters, n.thin = 5, progress.bar="none")

   # Compute DIC
   DIC    <- dic.samples(model,n.iter=iters,n.thin = 5, progress.bar="none")

   # Compute WAIC
   like   <- rbind(samps[[1]],samps[[2]]) # Combine samples from the two chains
   fbar   <- colMeans(like)
   Pw     <- sum(base::apply(log(like),2,var))
   WAIC   <- -2*sum(log(fbar))+2*Pw

   DIC
```

```{r}
   WAIC;Pw
```

\section{Model 2: Gaussian random effects}
```{r}
mod <- textConnection("model{
     for(i in 1:n){
       Y[i]          ~ dbern(pi[i])
       logit(pi[i]) <- beta[1]        + X[i,1]*beta[2] + X[i,2]*beta[3] + 
                       X[i,3]*beta[4] + X[i,4]*beta[5] + X[i,5]*beta[6] + 
                       theta[village[i]]
       like[i]      <- dbin(Y[i],pi[i],1) # For WAIC computation
     }
     for(j in 1:6){beta[j] ~ dnorm(0,0.01)}
     for(j in 1:65){theta[j] ~ dnorm(0,tau)}
     tau   ~ dgamma(0.1,0.1)
   }")

   data   <- list(Y=Y,X=X,n=n,village=village)
   model  <- jags.model(mod,data = data, n.chains=chains,quiet=TRUE)
   update(model, burn, progress.bar="none")
   samps  <- coda.samples(model, variable.names=c("like"), 
                          n.iter=iters, n.thin = 5,progress.bar="none")

   # Compute DIC
   DIC    <- dic.samples(model,n.iter=iters,n.thin = 5,progress.bar="none")

   # Compute WAIC
   like   <- rbind(samps[[1]],samps[[2]])
   fbar   <- colMeans(like)
   Pw     <- sum(base::apply(log(like),2,var))
   WAIC   <- -2*sum(log(fbar))+2*Pw

   DIC

```
```{r}
   WAIC;Pw
```

\section{Model 3: Double-exponential random effects}

```{r}
mod <- textConnection("model{
     for(i in 1:n){
       Y[i]          ~ dbern(pi[i])
       logit(pi[i]) <- beta[1]        + X[i,1]*beta[2] + X[i,2]*beta[3] + 
                       X[i,3]*beta[4] + X[i,4]*beta[5] + X[i,5]*beta[6] + 
                       theta[village[i]]
       like[i]      <- dbin(Y[i],pi[i],1) # For WAIC computation
     }
     for(j in 1:6){beta[j] ~ dnorm(0,0.01)}
     for(j in 1:65){theta[j] ~ ddexp(0,tau)}
     tau   ~ dgamma(0.1,0.1)
   }")

   data   <- list(Y=Y,X=X,n=n,village=village)
   model  <- jags.model(mod,data = data, n.chains=chains,quiet=TRUE)
   update(model, burn, progress.bar="none")
   samps  <- coda.samples(model, variable.names=c("like"), 
                          n.iter=iters, n.thin = 5,progress.bar="none")

   # Compute DIC
   DIC    <- dic.samples(model,n.iter=iters,n.thin = 5,progress.bar="none")

   # Compute WAIC
   like   <- rbind(samps[[1]],samps[[2]])
   fbar   <- colMeans(like)
   Pw     <- sum(base::apply(log(like),2,var))
   WAIC   <- -2*sum(log(fbar))+2*Pw

   DIC
```

```{r}
   WAIC;Pw
```

\textbf{Summary:} Both WAIC and DIC show strong support for including village random effects but cannot distinguish between Gaussian and double-exponential random effect distributions.