---
title:
output: pdf_document
---

\section{Spatial modeling of gun-related homicide rates}

\subsection{Linear models with correlated data}

The data for this analysis come from “Firearm legislation and firearm mortality in the USA: a cross-sectional, state-level study” by Kalesan et. al. (2016). The response variable, $Y_i$, is the log firearm-related death rate (i.e., the log of the number of deaths divided by the population) in 2010 in state $i$. This is regressed onto five potential confounders,

1.log 2009 firearm death rate per 10,000 people
2. Firearm ownership rate quartile
3. Unemployment rate quartile
4. Non-firearm homicide rate quartile
5. Firearm export rate quartile

The covariate of interest is the number of gun control laws in effect in the state. This gives $p=6$ covariates.

We fit the linear model

$$Y_i=\beta_0+\sum_{j=1}^pX_i\beta_j + \varepsilon_i.$$



We compare the usual non-spatial model with $\varepsilon_i\sim\mbox{Normal}(0,\sigma^2)$  with the spaital model $\mbox{Cov}(\varepsilon_1,…,\varepsilon_n)\sim\mbox{Normal}(0,\Sigma)$. The covariance

$$\Sigma=\tau^2S+ \sigma^2I_n$$

is decomposed into a spatial covariance $\tau^2 S$ and a non-spatial covariance $\sigma^2 I_n$. The spatial covariance follows the conditionally-autoregressive model 
$S = (M-\rho A)^{-1}$ wheree $A$ is the adjacency matrix with $(i,j)$ element is equal to 1 if states $i$ and $j$ are neighbors and zero otherwise, and $M$is the diagonal matrix with $i^{th}$ diagonal element equal to the number of states that neighbor state $i$.

\section{Load the data}

```{r,echo=FALSE}
set.seed(0820)
load("/Users/kamaladadashova/Documents/Ph.D. Courses/APPLIED BAYESIAN /Bayesian lecture/Week 6 Linear Models/guns.RData")
```

```{r}
 Y     <- log(10000*Y/N)
 Z[,1] <- log(Z[,1])
 X     <- cbind(1,Z,rowSums(X))

 # Remove AK and HI
 Y <- Y[-c(2,11)]
 X <- X[-c(2,11),]

 n <- length(Y)
 p <- ncol(X)
```


\section{Fit the non-spatial model}

```{r}
library(rjags)
ns_model <- "model{

   # Likelihood
   for(i in 1:n){
      Y[i]   ~ dnorm(mu[i],taue)
      mu[i] <- inprod(X[i,],beta[])
   }
   # Priors
   for(j in 1:p){beta[j] ~ dnorm(0,0.01)}
   taue ~ dgamma(0.1,0.1)
   sig <- 1/sqrt(taue)

 }"

 library(rjags)
 dat    <- list(Y=Y,n=n,X=X,p=p)
 init   <- list(beta=rep(0,p))
 model1 <- jags.model(textConnection(ns_model),
                      inits=init,data = dat,quiet=TRUE)
 update(model1, 10000, progress.bar="none")
 samp1   <- coda.samples(model1, 
            variable.names=c("beta","sig"), 
            n.iter=20000, progress.bar="none")
 summary(samp1)
```


\section{Create an adjacency matrix for the states in the US}

```{r}
 library(maps)
library(sf)
library(spdep)
library(rmapshaper)

# Create the USA state map
usa.state <- map(database = "state", fill = TRUE, plot = FALSE)

# Convert to an sf object
usa.sf <- st_as_sf(map("state", plot = FALSE, fill = TRUE))

# Clean the geometries to fix any issues
usa.sf <- st_make_valid(usa.sf)

# If there are still issues, simplify the geometries
usa.sf <- ms_simplify(usa.sf, keep_shapes = TRUE)

# Create neighborhood structure
usa.nb <- poly2nb(usa.sf)

# Convert to adjacency matrix
A <- nb2mat(usa.nb, style = "B")

# Remove DC (8th row/column)
A <- A[-8, ]
A <- A[, -8]

# Create the diagonal matrix
M <- diag(rowSums(A))




```

\section{Fit the spatial model}

```{r}
sp_model <- "model{

   # Likelihood
   for(i in 1:n){
      Y[i]  ~ dnorm(mu[i]+S[i],taue)
   }
   S[1:n] ~ dmnorm(zero[1:n],taus*Omega[1:n,1:n])
   for(i in 1:n){
      mu[i]   <- inprod(X[i,],beta[])
      zero[i] <- 0
   }
   Omega[1:n,1:n]<-M[1:n,1:n]-rho*A[1:n,1:n]

   # Priors
   for(j in 1:p){beta[j] ~ dnorm(0,0.01)}
   taue ~ dgamma(0.1,0.1)
   taus ~ dgamma(0.1,0.1)
   rho  ~ dunif(0,1)
   sig[1] <- 1/sqrt(taue)
   sig[2] <- 1/sqrt(taus)
  }"


  library(rjags)
  dat    <- list(Y=Y,n=n,X=X,A=A,M=M,p=p)
  init   <- list(rho=0.99,beta=lm(Y~X-1)$coef)
  model2 <- jags.model(textConnection(sp_model),
                       inits=init,data = dat,quiet=TRUE)
  update(model2, 10000, progress.bar="none")
  samp2  <- coda.samples(model2, 
            variable.names=c("beta","rho","sig"), 
            n.iter=20000, progress.bar="none")

  summary(samp2)
```

```{r}
  rho <- samp2[[1]][,8]
  hist(rho,breaks=100)
```

Summary: The spatial dependence parameter is estimated to be near one, indicating strong spatial dependence.

\section{Compare the results across models}

The objective is to determine if the coefficient corresponding to the number of gun laws $\beta_7$ s non-zero. Below we compare its posterior distribution for the spatial and non-spatial models.

```{r}
b1  <- samp1[[1]][,7]
 b2  <- samp2[[1]][,7]
 r   <- c(-0.035,0.015)
 d1  <- density(b1,from=r[1],to=r[2])
 d2  <- density(b2,from=r[1],to=r[2])

 plot(NA,xlim=r,ylim=c(0,max(d1$y)),
         xlab="Beta",ylab="Posterior density")
 lines(d1$x,d1$y)
 lines(d2$x,d2$y,lty=2)
 legend("topright",c("Non-spatial","Spatial"),lty=1:2,bty="n",cex=1.5)
 
```


```{r}
 mean(b1<0)
 mean(b2<0)
```

Summary: Both models provide evidence of a negative relationship between the number of gun laws and the firearm-related death rate. However, there is more uncertainty in the spatial model, which is likely more realistic.

