---
title: 
output: pdf_document
---
\section{Analysis of the 2016 US Presidential Election}


he data for this analysis come from Tony McGovern. The response variable, $Y_i$
, is the percentage change in Republican (GOP) support from 2012 to 2016, i.e.,

$$100\left(\frac{\mbox{\% in 2016}}{\mbox{\% in 2012}}-1\right),$$

in county $i=1,…,n$. 

The $p=10$ covariates $X_{ij}$ are county-level census variables obtained from Kaggle are:


Population, percent change - April 1, 2010 to July 1, 2014

Persons 65 years and over, percent, 2014

Black or African American alone, percent, 2014

Hispanic or Latino, percent, 2014

High school graduate or higher, percent of persons age 25+, 2009-2013

Bachelor's degree or higher, percent of persons age 25+, 2009-2013

Homeownership rate, 2009-2013

Median value of owner-occupied housing units, 2009-2013

Median household income, 2009-2013

Persons below poverty level, percent, 2009-2013


For a county in state $s$, we assume the linear model

$$Y_i=\beta_{0s}+\sum_{j=1}^pX_i\beta_{sj} + \varepsilon_i,$$

where $\beta_{js}$ is the effect of covariate $j$ in state $s$. We compare three models for the $\beta_{js}$. 

1. Constant slopes: $\beta_{js}=\beta_j$ for all counties.

2. Varying slopes with uninformative priors: $\beta_{js} \sim\mbox{Normal}(0,100)$

3. Varying slopes with informative priors: $\beta_{js} \sim\mbox{Normal}(\mu_j,\sigma_j^2)$.

In the third model, the means $\mu_j$ and variances $\sigma_j^2$ are given prior and estimated from the data, therefore information is pooled across states via the prior. The three methods are compared using DIC, and final results are compared across models.

```{r}
library(choroplethr)
load("/election_2008_2016.RData")
X     <- scale(X)    # Standardize the covariates
 X     <- cbind(1,X)
 short <- c("Intercept", "Pop change", "65+","African American",
            "Hispanic","HS grad","Bachelor's",
            "Homeownership rate","Home value",
            "Median income", "Poverty")
 names <- c("Intercept", as.character(names[1:11,2]))
 colnames(X) <- short

 # Define a function to make county maps
 county_plot <- function(fips,Y,main="",units=""){
   library(choroplethr)
   temp  <- as.data.frame(list(region=fips,value=Y))
   suppressWarnings(county_choropleth(temp,title=main,legend=units))
 }
 county_plot(fips,Y,main="Percent increase in GOP support",units="")
```


```{r}
 county_plot(fips,X[,7],main=names[7],units="")
```


```{r}
# Remove AK, HI and DC due to missing data
 state <- as.character(all_dat[,3])
 AKHI  <- state=="AK" | state=="HI" | state=="DC"
 fips  <- fips[!AKHI]
 Y     <- Y[!AKHI]
 X     <- X[!AKHI,]
 state <- state[!AKHI]

 # Assign a numeric id to the counties in each state
 st    <- unique(state)
 id    <- rep(NA,length(Y))
 for(j in 1:48){
   id[state==st[j]]<-j
 }

 n <- length(Y) # number of counties
 N <- 48        # number of states
 p <- ncol(X)   # number of covariates

 set.seed(0820)
 iters <- 50000
 burn  <- 10000
```

\section{Constant slopes}

```{r}
model1_string <- "model{

   # Likelihood
   for(i in 1:n){
      Y[i]   ~ dnorm(mu[i],taue)
      mu[i] <- inprod(X[i,],beta[])
   }
   # Priors
   for(j in 1:p){beta[j] ~ dnorm(0,0.01)}
   taue ~ dgamma(0.1,0.1)
   sig <- 1/sqrt(taue)

   # WAIC calculations
   for(i in 1:n){
     like[i]    <- dnorm(Y[i],mu[i],taue)
   }
 }"

 library(rjags)

 # Load the model
 dat    <- list(Y=Y,n=n,X=X,p=p)
 init   <- list(beta=rep(0,p))
 model1 <- jags.model(textConnection(model1_string),n.chains=2,
                      inits=init,data = dat,quiet=TRUE)
 # Generate samples
 update(model1, burn, progress.bar="none")
 samp1   <- coda.samples(model1, 
            variable.names="beta", 
            n.iter=iters, progress.bar="none")

 # Compile results
 ESS1    <- effectiveSize(samp1)
 out1    <- summary(samp1)$quantiles
 rownames(out1)<-short

 # Compute DIC
 dic1    <- dic.samples(model1,n.iter=iters,progress.bar="none")

 # Compute WAIC
 waic1   <- coda.samples(model1, 
            variable.names=c("like"), 
            n.iter=iters, progress.bar="none")
 like1   <- waic1[[1]]
 fbar1   <- colMeans(like1)
 P1      <- sum(base::apply(log(like1),2,var))
 WAIC1   <- -2*sum(log(fbar1))+2*P1
```

\section{Slopes as fixed effects}

```{r}
model2_string <- "model{

   # Likelihood
   for(i in 1:n){
      Y[i]    ~ dnorm(mnY[i],taue)
      mnY[i] <- inprod(X[i,],beta[id[i],])
   }

   # Slopes
   for(j in 1:p){for(i in 1:N){
       beta[i,j] ~ dnorm(0,0.01)
   }}

   # Priors
   taue ~ dgamma(0.1,0.1)


   # WAIC calculations
   for(i in 1:n){
     like[i]    <- dnorm(Y[i],mnY[i],taue)
   }
  }"


  library(rjags)

  # Load the model
  dat    <- list(Y=Y,n=n,N=N,X=X,p=p,id=id)
  init   <- list(beta=matrix(0,N,p))
  model2 <- jags.model(textConnection(model2_string),n.chains=2,
                       inits=init,data = dat,quiet=TRUE)

  # Generate samples
  update(model2, burn, progress.bar="none")
  samp2  <- coda.samples(model2, 
            variable.names="beta", 
            n.iter=iters, progress.bar="none")

  # Compile results  
  ESS2     <- effectiveSize(samp2)
  sum      <- summary(samp2)$stat
  post_mn2 <- matrix(sum[,1],N,p)
  post_sd2 <- matrix(sum[,2],N,p)

  # Compute DIC
  dic2   <- dic.samples(model2,n.iter=iters,progress.bar="none")

  # Compute WAIC
  waic2   <- coda.samples(model2, 
             variable.names=c("like"), 
             n.iter=iters, progress.bar="none")
  like2   <- waic2[[1]]
  fbar2   <- colMeans(like2)
  P2      <- sum(base::apply(log(like2),2,var))
  WAIC2   <- -2*sum(log(fbar2))+2*P2
```


\section{Slopes as random effects}

```{r}
 model3_string <- "model{

   # Likelihood
   for(i in 1:n){
      Y[i]    ~ dnorm(mnY[i],taue)
      mnY[i] <- inprod(X[i,],beta[id[i],])
   }

   # Random slopes
   for(j in 1:p){
     for(i in 1:N){
       beta[i,j] ~ dnorm(mu[j],taub[j])
     }
     mu[j]    ~ dnorm(0,0.01)
     taub[j]  ~ dgamma(0.1,0.1)
   }

   # Priors
   taue ~ dgamma(0.1,0.1)


   # WAIC calculations
   for(i in 1:n){
     like[i]    <- dnorm(Y[i],mnY[i],taue)
   }
  }"


  library(rjags)

  # Load the model
  dat    <- list(Y=Y,n=n,N=N,X=X,p=p,id=id)
  init   <- list(beta=matrix(0,N,p))
  model3 <- jags.model(textConnection(model3_string),n.chains=2,
                       inits=init,data = dat,quiet=TRUE)

  # Generate samples
  update(model3, burn, progress.bar="none")
  samp3  <- coda.samples(model3, 
            variable.names="beta", 
            n.iter=iters, progress.bar="none")

  # Compile results
  ESS3     <- effectiveSize(samp3)
  sum      <- summary(samp3)$stat
  post_mn3 <- matrix(sum[,1],N,p)
  post_sd3 <- matrix(sum[,2],N,p)

  # Compute DIC
  dic3    <- dic.samples(model3,n.iter=iters,progress.bar="none")

  # Compute WAIC
  waic3   <- coda.samples(model3, 
             variable.names=c("like"), 
             n.iter=iters, progress.bar="none")
  like3   <- waic3[[1]]
  fbar3   <- colMeans(like3)
  P3      <- sum(base::apply(log(like3),2,var))
  WAIC3   <- -2*sum(log(fbar3))+2*P3
```

\section{Check convergence}

```{r}
 ESS1
```
```{r}
 hist(ESS2)
```

```{r}
 hist(ESS3)
```

Summary: The effective sample size is large for all parameters and all models, therefore it seems the MCMC algorithm has converged.

\section{Summarize the non-spatial model}

```{r}
library(kableExtra)
 kbl(round(out1,2))
```

Summary: All but one (home ownership rate) of the covariates have 95% interval that excludes zero. GOP support tended to increase in counties with

decreasing population
high proportion of seniors and high school graduates
low proportions of African Americans and Hispanics
High income but low home value
High poverty rate

\section{Compare models with DIC}

```{r}
dic1
dic2
dic3
```

Summary: The first model with constant slopes is the simplest but fits the data poorly and thus has highest DIC. The second model with different slopes in each state has the best fit (smallest mean deviance), but is too complicated and has large
$p_D$ The final model has fairly small mean deviance and $p_D$,  and thus balances fit and complexity to give the smallest DIC.


\section{Compare models with WAIC}

```{r}
WAIC1; P1
WAIC2; P2
WAIC3; P3
```

Summary: WAIC agrees with DIC. Both prefer Model 3 with the regression coefficients treated as random effects.

\section{Explore the results of the final model}

```{r}
boxplot(post_mn3)
```

Summary: The effect of the proportion of college graduates varies the most across states

\section{Explore the three estimate of the effects of college graduates}

```{r}
 # Posterior mean
  county_plot(fips,post_mn3[id,7],
              main="Proportion of college graduates - posterior mean")
```

```{r}
 # Posterior sd
  county_plot(fips,post_sd3[id,7],
  main="Proportion of college graduates - posterior SD")
```


Summary: The effect of the proportion of college graduates is the strongest (most negative) in the midwest.
